Gaussian Processes Autoregressive Models for the Disturbance Storm Time Index

### 2016-05-02 swsc.tex
Updates from ShareLaTeX

The distribution of $\mathbf{f_*}| \mathbf{X},\mathbf{y},\mathbf{X_*}$ is known in Bayesian analysis research as the \emph{Posterior Predictive Distribution}. This illustrates a key difference between \emph{Gaussian Processes} and other regression models such as \emph{Neural Networks}, \emph{Linear Models} and \emph{Support Vector Machines}, that a \emph{Gaussian Process} model does not generate point predictions for new data but outputs a predictive distribution for the quantity sought, this allows us to construct error bars on the predictions. This property of Bayesian models such as \emph{Gaussian Processes} makes them suitable for Space Weather forecasting applications.

@@

Figure \ref{fig:timingErr} compares the time discrepancy of predicting the storm peak, again as seen above the \emph{Persistence} model will always have $\Delta t_{peak} = -1$. Two central themes come out of these charts.

\begin{enumerate}
    \item Persistence Behavior: It is clear that the \emph{persistence} behavior in the $D_{st}$ values is very strong i.e. the trivial predictive model $\hat{D_{st}}(t) = D_{st}(t-1)$ gives excellent performance. In fact it can be seen in these figures that the \emph{Persistence} model is outperforms every model compared by \citet{Ji2012} in the context of OSA prediction of $D_{st}$. Therefore any model that is proposed to tackle the OSA prediction problem for $D_{st}$ must be compared to and show visible gains over the \emph{Persistence} model.
    \item Assessment of Gaussian Process Models: The results indicate that there are many benefits to be gained by the application of Gaussian Process models for prediction of $D_{st}$. The \emph{GP-AR} and \emph{GP-ARX} not only outperform the models compared in \citet{Ji2012} but also the \emph{Persistence} model. 
\end{enumerate}

@@

We conclude this paper with three key conclusions/insights learned from our research so far.
   \begin{enumerate}
      \item \emph{Persistence} model must be central in the model evaluation process in the context of \emph{One Step Ahead} prediction of the $D_{st}$ index. This is due to its excellent performance in terms of errors and correlations. 
      \item \emph{Gaussian Process} AR and ARX models give encouraging benefits in OSA prediction even when compared to the \emph{Persistence} model. Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data.

>>>>

The distribution of $\mathbf{f_*}| \mathbf{X},\mathbf{y},\mathbf{X_*}$ is known in Bayesian analysis as the \emph{Posterior Predictive Distribution}. This illustrates a key difference between \emph{Gaussian Processes} and other regression models such as \emph{Neural Networks}, \emph{Linear Models} and \emph{Support Vector Machines}, that a \emph{Gaussian Process} model does not generate point predictions for new data but outputs a predictive distribution for the quantity sought, this allows us to construct error bars on the predictions. This property of Bayesian models such as \emph{Gaussian Processes} makes them suitable for Space Weather forecasting applications.

@@

Figure \ref{fig:timingErr} compares the time discrepancy of predicting the storm peak, again as observed above the \emph{Persistence} model will always have $\Delta t_{peak} = -1$, the GP-AR and GP-ARX models give better performance than the other models in terms of timing error in prediction of storm peaks.

In figure \ref{fig:predictions} we show the OSA predictions generated by the TL, NM and GP-ARX models for one event from the list of storms in \citet{Ji2012}. This storm is interesting due to its strength ($Dst_{min} = -289 nT$) and the presence of two distinct peaks. In this particular event the GP-ARX model approximates the twin peaks as well as the onset and decay phases quite faithfully which is lacking in the NM and TL model predictions. The TL model recognises the presence of twin peaks but over estimates the second one as well as the decay phase, while the NM model is much delayed in the prediction of the initial peak and fails to approximate the time and intensity of the second peak.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_pred.png}
      \caption{Comparison of OSA predictions generated by the NM, TL and GP-ARX models for the storm event $9^{th}$ November $2004$ 11:00 UTC - $11^{th}$ November $2004$ 09:00 UTC}
         \label{fig:predictions}
   \end{figure}

\begin{comment}
\begin{enumerate}
    \item Persistence Behavior: It is clear that the \emph{persistence} behavior in the $D_{st}$ values is very strong i.e. the trivial predictive model $\hat{D_{st}}(t) = D_{st}(t-1)$ gives excellent performance. In fact it can be seen in these figures that the \emph{Persistence} model is outperforms every model compared by \citet{Ji2012} in the context of OSA prediction of $D_{st}$. Therefore any model that is proposed to tackle the OSA prediction problem for $D_{st}$ must be compared to and show visible gains over the \emph{Persistence} model.
    \item Assessment of Gaussian Process Models: The results indicate that there are many benefits to be gained by the application of Gaussian Process models for prediction of $D_{st}$. The \emph{GP-AR} and \emph{GP-ARX} not only outperform the models compared in \citet{Ji2012} but also the \emph{Persistence} model. 
\end{enumerate}
\end{comment}

@@

We conclude this paper with three key conclusions/insights learned from our research so far.
   \begin{enumerate}
      \item \emph{Persistence} model must be central in the model evaluation process in the context of \emph{One Step Ahead} prediction of the $D_{st}$ index. It is clear that the \emph{persistence} behavior in the $D_{st}$ values is very strong i.e. the trivial predictive model $\hat{D_{st}}(t) = D_{st}(t-1)$ gives excellent performance. In fact it can be seen in \ref{sec:res} that the \emph{Persistence} model is outperforms every model compared by \citet{Ji2012} in the context of OSA prediction of $D_{st}$. Therefore any model that is proposed to tackle the OSA prediction problem for $D_{st}$ must be compared to the \emph{Persistence} model and show visible gains above it.

      \item \emph{Gaussian Process} AR and ARX models give encouraging benefits in OSA prediction even when compared to the \emph{Persistence} model. Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data.
